---
title: "Reproducing SD model calibration with Tidyverse and deSolve"
author: |
  | Jair Andrade, MSc.
  | National University of Ireland Galway
date: "April 2019"
output:
  pdf_document: default
  html_document:
    df_print: paged
header-includes: \usepackage{amsmath}
                 \usepackage[utf8]{inputenc}
---

```{r include = FALSE}
library(knitr)
opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, fig.align = "center", cache = TRUE)
```

### Introduction

Demonstrate how to reproduce model calibration with 'tidyverse' and 'deSolve'

"we hope this primer provides aspiring researches with the stepping stones for building new techniques that further expand the analytical toolbox of dynamic modelers"

### Model

It is used an auxiliary function to returns a SD model that embeds exogenous variables CO & SC. In doing so, these two variables are not required as inputs for subsequent simulations. The model take as parameters the effect of work pressure on time per order, a time constant for decreasing time per order, a time constant for increasing time per order and the initial desired time per order. Thanks to the deSolve package[cite required], it is possible to simulate the SD model and return the actual time per order's behaviour over time.

```{r model}
library(deSolve)
library(tidyverse)
# alpha: effect of work pressure on time per order
# tau.td: time constant for decreasing time per order
# tau.ti: time constant for increasing time per order
# TPOD0: the initial desired time per order
# CO: Incoming Orders Ot
# SC: Service Capacity Ct

generate_model <- function(CO, SC){
  os_model <- function(params) {
    alpha <- params[[1]]
    tau_td <- params[[2]]
    tau_ti <- params[[3]]
    TPOD0 <- params[[4]]
    STEP <- 1 / 64
    
    # Create time vector
    simtime <- seq(53, 104, by = STEP)
    
    # Create stocks vector, with initial values
    stocks  <- c(sDesired.Time.Order = TPOD0) # Desired time per order initialised
    
    # Create auxiliaries vector, with values
    auxs    <- c(aElasticity.Work.Pressure = alpha,
                 aQuality.Pressure = 1,
                 aDecreasing.Time.Order = tau_td,
                 aIncreasing.Time.Order = tau_ti,
                 aMinimum.Order.Processing.Time = 0.1)
    
    orders <- approxfun(x = seq(53, 104),
                        y = CO,
                        method = "linear",
                        yleft = min(CO),
                        yright = max(CO))
    
    service.capacity <- approxfun(x = seq(53, 104),
                                  y = SC,
                                  method = "linear",
                                  yleft = min(SC),
                                  yright = max(SC))
    
    equations <- function(time, stocks, auxs){
      with(as.list(c(stocks, auxs)),{
        aOrders <- orders(time)
        aDesired.Service.Capacity <- sDesired.Time.Order * aOrders
        aActual.Service.Capacity <- service.capacity(time)
        aRatio.SC <- aDesired.Service.Capacity / aActual.Service.Capacity
        aWork.Pressure <- ifelse(aDesired.Service.Capacity == 0,
                                 1000000,
                                 (aDesired.Service.Capacity / aActual.Service.Capacity) ^ aElasticity.Work.Pressure)
        aTime.Per.Order <- max(aMinimum.Order.Processing.Time,
                               aQuality.Pressure * aWork.Pressure * sDesired.Time.Order)
        # DTO = Desired time per order
        aTime.Adjust.DTO <- ifelse(aTime.Per.Order > sDesired.Time.Order,
                                   aIncreasing.Time.Order, aDecreasing.Time.Order)
        fChange.DTO <- (aTime.Per.Order - sDesired.Time.Order) / aTime.Adjust.DTO
        d_sDesired.Time.Order_dt <- fChange.DTO
        return (list(c(d_sDesired.Time.Order_dt),
                     TPO = aTime.Per.Order))
      })
    }
    # Call Solver, and store results in a data frame
    output <- data.frame(ode(y = stocks, times = simtime, func = equations,
                             parms = auxs, method = "euler"))
    result <- output %>% filter(time - round(time) == 0) %>% select(time, TPO)
  }
  os_model
}

reg.data <-read_csv('Data/ServiceQualityData2.csv')
names(reg.data) <- make.names(names(reg.data))
n.rows  <-nrow(reg.data)
weeks   <- reg.data %>% pull(Week)
ACT.TPO <-reg.data %>% pull(ACT.TPO)  # actual time per order
CO      <-reg.data %>% pull(CO)  # incomining orders Ot
SC     <- reg.data %>% pull(SC)  # actual service capacity Ct
os_model <- generate_model(CO, SC)
```

To test the model, it is simulated using parameters estimated by other people

```{r testModel}
param_set <- data.frame(stringsAsFactors = FALSE,
  source = c("Oliva & Sterman (2001)", "Oliva & Sterman (2001)",
    "Oliva & Sterman (2001)", "Oliva & Sterman (2001)",
    "Dogan(2007)", "Dogan(2007)", "Dogan(2007)", "Dogan(2007)",
    "Oliva (1996)", "Oliva (1996)", "Oliva (1996)", "Oliva (1996)"),
  variable = c("alpha", "tau_td", "tau_ti", "TPOD0", "alpha", "tau_td",
    "tau_ti", "TPOD0", "alpha", "tau_td", "tau_ti", "TPOD0"),
  value = c(-0.64, 18.83, 8140000, 1.08, -0.6, 12.48, 85600, 1.07,
    -0.6393, 18.7321, 814000, 1.0758)
)
run_model <- function(par_source){
  params <- param_set %>% filter(source == par_source) %>% pull(value)
  result <- os_model(params) %>% mutate(source = par_source)
}

sources <- unique(param_set$source)
sim_df  <- data.frame(time = 53:104, TPO = ACT.TPO, source = 'actual data') %>% 
  bind_rows(map_df(sources, run_model))

g <- ggplot(sim_df, aes(x = time, y = TPO, group = source)) +
  geom_line(aes(colour = source), size = 1) +
  scale_colour_manual(values = c('#525252', '#969696','#cccccc', '#f7f7f7')) +
  theme_classic() +
  labs(x = "Time", y = "Time per order")

print(g)
```

